---
title: "Normality Test"
output: html_document
---

### Set Up

```{r cars}
#First check that package required is installed, if not install it
# Specify your packages
needed_packages <- c("pastecs", "ggplot2", "semTools", "FSA")                                    
# Extract not installed packages
not_installed <- needed_packages[!(needed_packages %in% installed.packages()[ ,"Package"])]    
# Install not installed packages
if(length(not_installed)) install.packages(not_installed)                              

library(pastecs) #For creating descriptive statistic summaries
library(ggplot2) #For creating histograms with more detail than plot
library(semTools) #For skewness and kurtosis
library(FSA) #calculate the percentage that are within a range 
```

### Load Data

```{r}
#Read Math dataset from CSV
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
#Merge both datasets
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
```
### Normality Test for Math Grade-1

##### Plot histogram

```{r}
#Analyse the normale distribution of Grade
gg<-ggplot(data = d3,mapping = aes(x=d3$G1.x))
#Change the label of the x axis
gg <- gg + labs(x="Math Grade-1")

#manage binwidth and colours
gg <- gg + geom_histogram(binwidth=2, colour="black", aes(y=..density.., fill=..count..))
gg <- gg + scale_fill_gradient("Count", low="#EBF5FB", high="#1B4F72")

#adding a normal curve
#use stat_function to compute a normalised score for each value of tpcoiss
#pass the mean and standard deviation
#use the na.rm parameter to say how missing values are handled
gg <- gg + stat_function(fun=dnorm, color="red",args=list(mean=mean(d3$G1.x, na.rm=TRUE), sd=sd(d3$G1.x, na.rm=TRUE)))
gg
```

##### Plot Q-Q Plot
```{r}
#Create a qqplot
qqnorm(d3$G1.x)
qqline(d3$G1.x, col=2) #show a line on theplot
```

##### Summary Statistics

```{r}
#stat.desc is a function from pastecs - make sure you include the basic switch=F to ensure you don't get scienfitic notation
stat.desc(d3$G1.x, basic=F)
```

##### Skew and kurtosis Test:
The standardized scores(value/std. error) for Skewness and kurtosis between +/- 2 are considered acceptable in order to prove **normal univariate distribution**.

```{r}
#We can make our decision based on the value of the standardised score for skew and kurtosis
#We divide the skew statistic by the standard error to get the standardised score
#This will indicate if we have a problem
tpskew<-skew(d3$G1.x)
tpskew[1]/tpskew[2]

tpkurt<-kurtosis(d3$G1.x)
tpkurt[1]/tpkurt[2]
```
As the standardized scores of the Skew and kurtosis is outside of +/- 2 the impact need to be accessed. We need to check how much percentage of standardized score falling outside of 1.96. If it is less than 5%, then it is safe to say as a ***Normal Distribution***.

As the number of sample is more than 80, we need to getback with the percentage of observation falling outside 3.29.

```{r}
#and by calculating the percentage of standardised scores for the variable itself that are outside our acceptable range
#This will tell us how big a problem we have
# Calculate the percentage of standardised scores that are greated than 1.96
# the perc function which is part of the FSA package which calculate the percentage that are within a range - you can look for greater than "gt", greater than or equal "geq", "gt", less than or equal "leq",  or less than "lt"),
# scale is a function that creates z scores, abs gets absolute value

ztpcoiss<- abs(scale(d3$G1.x))

perc(as.numeric(ztpcoiss), 1.96, "gt")
perc(as.numeric(ztpcoiss), 3.29, "gt")
```
### Normality Test for Portuguese Grade-1

##### Plot histogram

```{r}
#Analyse the normale distribution of Grade
gg<-ggplot(data = d3,mapping = aes(x=d3$G1.y))
#Change the label of the x axis
gg <- gg + labs(x="Portuguese Grade-1")

#manage binwidth and colours
gg <- gg + geom_histogram(binwidth=2, colour="black", aes(y=..density.., fill=..count..))
gg <- gg + scale_fill_gradient("Count", low="#EBF5FB", high="#1B4F72")

#adding a normal curve
#use stat_function to compute a normalised score for each value of tpcoiss
#pass the mean and standard deviation
#use the na.rm parameter to say how missing values are handled
gg <- gg + stat_function(fun=dnorm, color="red",args=list(mean=mean(d3$G1.y, na.rm=TRUE), sd=sd(d3$G1.y, na.rm=TRUE)))
gg
```

##### Plot Q-Q Plot
```{r}
#Create a qqplot
qqnorm(d3$G1.y)
qqline(d3$G1.y, col=2) #show a line on theplot
```

##### Summary Statistics

```{r}
#stat.desc is a function from pastecs - make sure you include the basic switch=F to ensure you don't get scienfitic notation
stat.desc(d3$G1.y, basic=F)
```

##### Skew and kurtosis Test:
The standardized scores(value/std. error) for Skewness and kurtosis between +/- 2 are considered acceptable in order to prove **normal univariate distribution**.

```{r}
#We can make our decision based on the value of the standardised score for skew and kurtosis
#We divide the skew statistic by the standard error to get the standardised score
#This will indicate if we have a problem
tpskew<-skew(d3$G1.y)
tpskew[1]/tpskew[2]

tpkurt<-kurtosis(d3$G1.y)
tpkurt[1]/tpkurt[2]
```
As the standardized scores of the Skew and kurtosis is outside of +/- 2 the impact need to be accessed. We need to check how much percentage of standardized score falling outside of 1.96. If it is less than 5%, then it is safe to say as a ***Normal Distribution***.

As the number of sample is more than 80, we need to getback with the percentage of observation falling outside 3.29.

```{r}
#and by calculating the percentage of standardised scores for the variable itself that are outside our acceptable range
#This will tell us how big a problem we have
# Calculate the percentage of standardised scores that are greated than 1.96
# the perc function which is part of the FSA package which calculate the percentage that are within a range - you can look for greater than "gt", greater than or equal "geq", "gt", less than or equal "leq",  or less than "lt"),
# scale is a function that creates z scores, abs gets absolute value

ztpcoiss<- abs(scale(d3$G1.y))

perc(as.numeric(ztpcoiss), 1.96, "gt")
perc(as.numeric(ztpcoiss), 3.29, "gt")
```