---
title: "Independent t-tests"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---


 
### SETUP
```{r}

needed_packages <- c("pastecs", "ggplot2", "psych", "semTools", "FSA", "car", "coin", "rstatix","effectsize")                      
# Extract not installed packages
not_installed <- needed_packages[!(needed_packages %in% installed.packages()[ , "Package"])]    
# Install not installed packages
if(length(not_installed)) install.packages(not_installed) 

library(pastecs) #For creating descriptive statistic summaries
library(ggplot2) #For creating histograms with more detail than plot
library(psych) # Some useful descriptive functions
library(semTools) #For skewness and kurtosis
library(FSA) #For percentage
library(car) # For Levene's test for homogeneity of variance 
library(effectsize) #To calculate effect size for t-test
```
### Load Data

```{r}
#Read Math dataset from CSV
d1=read.table("student-mat.csv",sep=";",header=TRUE)
d2=read.table("student-por.csv",sep=";",header=TRUE)
#Merge both datasets
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
```
### Normality Test for Math Grade-1

##### Plot histogram

```{r}
#Analyse the normale distribution of Grade
gg<-ggplot(data = d3,mapping = aes(x=d3$G1.x))
#Change the label of the x axis
gg <- gg + labs(x="Math Grade-1")

#manage binwidth and colours
gg <- gg + geom_histogram(binwidth=2, colour="black", aes(y=..density.., fill=..count..))
gg <- gg + scale_fill_gradient("Count", low="#EBF5FB", high="#1B4F72")

#adding a normal curve
#use stat_function to compute a normalised score for each value of tpcoiss
#pass the mean and standard deviation
#use the na.rm parameter to say how missing values are handled
gg <- gg + stat_function(fun=dnorm, color="red",args=list(mean=mean(d3$G1.x, na.rm=TRUE), sd=sd(d3$G1.x, na.rm=TRUE)))
gg
```

##### Plot Q-Q Plot
```{r}
#Create a qqplot
qqnorm(d3$G1.x)
qqline(d3$G1.x, col=2) #show a line on theplot
```

##### Summary Statistics

```{r}
#stat.desc is a function from pastecs - make sure you include the basic switch=F to ensure you don't get scienfitic notation
stat.desc(d3$G1.x, basic=F)
```

##### Skew and kurtosis Test:
The standardized scores(value/std. error) for Skewness and kurtosis between +/- 2 are considered acceptable in order to prove **normal univariate distribution**.

```{r}
#We can make our decision based on the value of the standardised score for skew and kurtosis
#We divide the skew statistic by the standard error to get the standardised score
#This will indicate if we have a problem
tpskew<-skew(d3$G1.x)
tpskew[1]/tpskew[2]

tpkurt<-kurtosis(d3$G1.x)
tpkurt[1]/tpkurt[2]
```
As the standardized scores of the Skew and kurtosis is outside of +/- 2 the impact need to be accessed. We need to check how much percentage of standardized score falling outside of 1.96. If it is less than 5%, then it is safe to say as a ***Normal Distribution***.

As the number of sample is more than 80, we need to getback with the percentage of observation falling outside 3.29.

```{r}
#and by calculating the percentage of standardised scores for the variable itself that are outside our acceptable range
#This will tell us how big a problem we have
# Calculate the percentage of standardised scores that are greated than 1.96
# the perc function which is part of the FSA package which calculate the percentage that are within a range - you can look for greater than "gt", greater than or equal "geq", "gt", less than or equal "leq",  or less than "lt"),
# scale is a function that creates z scores, abs gets absolute value

ztpcoiss<- abs(scale(d3$G1.x))

perc(as.numeric(ztpcoiss), 1.96, "gt")
perc(as.numeric(ztpcoiss), 3.29, "gt")
```

### Differences - Parametric Tests
#### Independent t-test

H0 (Null): There is no difference in Math Grade - 1 score for students who are from Gabriel Pereira School and those from Mousinho da Silveira school

Ha (Alternate): There is a difference in Math Grade - 1 score for students who are from Gabriel Pereira School and those from Mousinho da Silveira school


#### Describe Math Grade-1 by group (Gabriel Pereira vs Mousinho da Silveira - variable School)
#### Levene's test
Conduct Levene's test for homogeneity of variance in library car - the null hypothesis is that variances in groups are equal so to assume homogeneity we woudl expect probaility to not be statistically significant.

H0 (Null): Math Grade-1 population variance are equal for both schools Gabriel Pereira and Mousinho da Silveira

Ha (Alternate): Math Grade-1 population variance are not equal for both schools Gabriel Pereira and Mousinho da Silveira

If Pr(>F) < 0.05 then we reject the Hypothesis meaning the variances are not equal
```{r Describe}
#Get descriptive stastitics by group - output as a matrix
psych::describeBy(d3$G1.x, d3$school, mat=TRUE)
car::leveneTest(G1.x ~ school, data=d3)
```
Pr(>F) is your probability - in this case it is not statistically significant so we can assume homogeneity.

Conduct the t-test from package stats. In this case we can use the var.equal = TRUE option to specify equal variances and a pooled variance estimate.
```{r}
stats::t.test(G1.x~school,var.equal=TRUE,data=d3)
#No statistically significant difference was found

res <- stats::t.test(G1.x~school,var.equal=TRUE,data=d3)
#Calculate Cohen's d
#artithmetically
effcd=round((2*res$statistic)/sqrt(res$parameter),2)
effcd
#Using function from effectsize package
effectsize::t_to_d(t = res$statistic, res$parameter)


#Eta squared calculation
effes=round((res$statistic*res$statistic)/((res$statistic*res$statistic)+(res$parameter)),3)
effes


```
### Reporting the results with Cohen's d effect

```
"An independent-samples t-test was conducted to compare Math Grade - 1 score for students who are from Gabriel Pereira School and those from Mousinho da Silveira school.No Significant difference in the scores for Math Grade - 1 was found (M=10.9, SD= 3.34 for students from Gabriel Pereira School, M= 10.55, SD= 3.48 for students from Gabriel Pereira School), (t(`r res$parameter`)= `r round(res$statistic,3)`), p = `r round(res$p.value,2)`. Cohen's d also indicated a very small effect size (`r effcd`)."

```

### Reporting the results with eta squred effect

```
"An independent-samples t-test was conducted to compare Math Grade - 1 score for students who are from Gabriel Pereira School and those from Mousinho da Silveira school.No Significant difference in the scores for Math Grade - 1 was found (M=10.9, SD= 3.34 for students from Gabriel Pereira School, M= 10.55, SD= 3.48 for students from Gabriel Pereira School), (t(`r res$parameter`)= `r round(res$statistic,3)`), p = `r round(res$p.value,2)`. A very small effect size was also indicted by the eta squared value (`r effes`)."

```


